
\section{Dynamic Semantics}
\label{base:semantics}

\subsection{Reduction}

We express reduction rules using \emph{substructural operational semantics} \cite{Simmons12} which are based on \emph{multiset rewriting} \cite{CervesatoS09}. For example, the rule for $\terminate$ can be written as:
$$ \proc{c}{\tclose{c}} \otimes \proc{d}{\twait{c}{P}} \lolli \monad{\proc{d}{P}}. $$
Note that the rule is written using linear connectives, however, these should not be confused with connectives we used for types. Instead, a rule of the form $A_1 \tensor \ldots \tensor A_n \lolli \monad{B_1 \tensor \ldots \tensor B_m}$ means we can replace the resources $A_1, \ldots, A_n$ with $B_1, \ldots, B_m$. The curly braces $\braces \ldots$ indicates a monad which essentially forces the rules to be interpreted as a multiset rewriting rule. $\monad{\exists x. F}$ generates a fresh $a$ and proceeds with $subst a x F$, while $c = d$ performs a global identification of $c$ and $d$ in the configuration. The rest of the rules are given in \cref{base:reduction-relation}.

\begin{figure}[!ht]
  \centering
  \begin{align*}
    % Id
    \irb{id}     \hspace{1em} & : \proc{c}{\tfwd{c}{d}} \lolli \monad{c = d} \\
    % Cut
    \irb{cut}    \hspace{1em} & : \proc{c}{\tspawn{x}{P_x}{Q_x}}
        \lolli \monad{\exists a. \proc{a}{P_a} \otimes \proc{c}{Q_a}} \\
    % One
    \irb{one} \hspace{1em} & : \proc{c}{\tclose{c}} \otimes \proc{d}{\twait{c}{P}}
      \lolli \monad{\proc{d}{P}} \\
    % Tensor
    \irb{tensor} \hspace{1em} & : \proc{c}{\tsend{c}{x}{P_x}{Q}} \otimes \proc{e}{\trecv{x}{c}{R_x}} \\
      & \hspace{2em} \lolli \monad{ \exists a. \proc{a}{P_{a}} \otimes \proc{c}{Q} \otimes \proc{e}{R_{a}} } \\
    % Internal
    \irb{internal} \hspace{1em} & : \proc{c}{\tselect{c}{i}{P}} \otimes \proc{d}{\tcase{c}{\tbranches Q I}} \otimes i \in I \\
      & \hspace{2em} \lolli \monad{ \proc{c}{P} \otimes \proc{d}{Q_i} } \\
    % Lolli
    \irb{lolli} \hspace{1em} & : \proc{c}{\trecv{x}{c}{P_x}} \otimes \proc{d}{\tsend{c}{x}{Q_x}{R}} \\
      & \hspace{2em} \lolli \monad{ \exists a. \proc{c}{P_{a}} \otimes \proc{a}{Q_a} \otimes \proc{d}{R} } \\
    % External
    \irb{external} \hspace{1em} & : \proc{c}{\tcase{c}{\tbranches P I}} \otimes \proc{d}{\tselect c i Q} \otimes i \in I \\
      & \hspace{2em} \lolli \monad{ \proc{c}{P_i} \otimes \proc{d}{Q} } \\
    % Recursive
    \irb{rec} \hspace{1em} & : \proc{c}{\tapp {\parens{\trec p {\tvector y} P}} {\tvector z}}
        \lolli \monad{ \proc c {\subst {\trec p {\tvector y} P} p {\subst {\tvector z} {\tvector y} P}} }
  \end{align*}
  \caption{Reduction rules for process configurations}
  \label[rules]{base:reduction-relation}
\end{figure}

We say that $\config$ reduces to or steps to $\config'$ if one application of the above rules transforms $\config$ into $\config'$, and write $\steps{\config}{\config'}$. We denote the reflexive transitive closure of $\stepArrow$ by $\stepManyArrow$.

An important observation about reductions in typed configurations is that they are constrained to one subtree, where the tree structure is implicit in the typing judgment as discussed in \cref{chapter/session-types/configurations}. This observation leads to the following framing rule.

\begin{definition}[Root of Reduction] We say that channel $c$ is the root of $\steps{\config}{\config'}$ if $\proc c P \in \config$ is rewritten by the reduction, and either it is the only process to the left of $\lolli$ (rules $\irb{id}, \irb{cut}, \irb{rec}$), or it is the client.
\end{definition}

\begin{lemma}[Framing]
  \label{framing}
  If $\providesCtx \config \ctx$ and $\steps{\config}{\config'}$ then there exist $\config_1, \config_2, \config_2'$ such that $\config = (\config_1, \config_2)$, $\config' = (\config_1, \config_2')$, $\steps{\config_2}{\config_2'}$, and $\provides {\config_2} c A$ where $c$ is the root of $\steps{\config}{\config'}$. In addition, if $\provides {\config_2'} c A$, then $\providesCtx {\config'} \ctx$.
\end{lemma}

\begin{proof}
  Assume $\providesCtx \config \ctx$ and $\steps{\config}{\config'}$ with root $c$. We proceed by induction on $\providesCtx \config \ctx$.
  \begin{itemize}
    \item For the single channel case, assume $\ctx = d : A$. If $d = c$, then we pick $\config_1 = \emptyConfig$, $\config_2 = \config$, and $\config_2' = \config'$. The result follows immediately. Otherwise, we know $\proc c P \in \config$ was not part of the reduction, so we can apply the induction hypothesis. We add $\proc c P$ to the $\config_1$ we get from the induction hypothesis.
    \item For the multiple channel case, assume $\ctx = d_1 : A_1, \ldots, d_n : A_n$, $\config = \config^1, \ldots, \config^n$, and $\provides {\config^i} {d_i} {A_i}$. We know the reduction must work on only one of the $\config_i$, so we apply the induction hypothesis on that portion of the context. We add the rest to $\config_1$.
  \end{itemize}
\end{proof}


\subsection{Progress}
\label{base:progress-section}

In a conventional functional language, the progress theorem states that a well-typed expression either is a value or it takes a reduction step. We do not have values in a process calculus, but there is a corresponding notion we call being poised. Intuitively, a process is poised if it is waiting on its client. We introduce a new judgment $\poised{\proc c P}$ capturing this notion and define it in \cref{poised}.

\begin{rules}[poised]{Poised processes}
  \infer{\poised{\proc{c}{\tclose{c}}}}
   {}
  \and \infer{\poised{\proc{c}{\tsend{c}{d}{P_d}{Q}}}}
   {}
  \and \infer{\poised{\proc{c}{\trecv{x}{c}{Q_x}}}}
   {}
  \and \infer{\poised{\proc{c}{\tselect{c}{i}{P}}}}
   {}
  \and \infer{\poised{\proc{c}{\tcase{c}{\tbranches{Q}{I}}}}}
   {}
\end{rules}

We say that a process configuration $\config$ is poised if every process in $\config$ is poised.  We will need the following inversion lemma about well-typed poised processes to handle type equality:

\begin{lemma}[Inversion of Process Typing]
\label{inversion-poised-eq}
  If $\typeRecD \ctx \emptyset P c A$ and $\poised{\proc c P}$, then:
  \begin{itemize}
    \item If $A \typeeq \terminate$ then $P = \tclose c$.
    \item If $A \typeeq A_1 \tensor A_2$ then $P = \tsend c d {Q_d} {P'}$.
    \item If $A \typeeq \internals A I$ then $P = \tselect c x {P'}$ where $x \in I$.
    \item If $A \typeeq A_1 \lolli A_2$ then $P = \trecv d c {P'}$.
    \item If $A \typeeq \externals A I$ then $P = \tcase c {\tbranches {P'} J}$ where $I \subseteq J$.
  \end{itemize}
\end{lemma}

\begin{proof}
  The proof is by induction on the derivation of $\typeRecD \ctx \emptyset P c A$ where $\ctx$ and $A$ are free.

  \begin{description}
    \item[Case $\id, \cut, \terminate\Left, \tensor\Left, \internal\Left, \lolli\Left, \external\Left, \rec :$] Impossible since $P$ is poised.

    \item[Case $\terminate\Right, \tensor\Right, \internal\Right, \lolli\Right, \external\Right :$] If the rule matches the expected type (e.g.\ $A \typeeq 1$ and the rule is $\terminate\Right$), then $P$ has the expected form and we are done. Otherwise, we use inversion on the type equality judgment to show that the case is impossible.

    \item[Case $\typeeq\Left :$] Follows immediately from the induction hypothesis.

    \item[Case $\typeeq\Right :$] Follows from the induction hypothesis and the fact that $\typeeq$ is transitive (\cref{eq-is-equivalence}).
  \end{description}
\end{proof}

We are now ready to state the progress theorem.

\begin{theorem}[Progress]
\label{base:progress}
If $\providesCtx \config \ctx$ then either
\begin{enumerate}
  \item $\steps{\config}{\config'}$ for some $\config'$, or
  \item $\config$ is poised.
\end{enumerate}
\end{theorem}

\begin{proof}
The proof is by induction on configuration typing followed by a nested induction on the typing derivation in the single channel case.

\begin{itemize}
  \item The case for multiple channels is simpler, so we will do that first. Assume $\config = \config_1, \ldots, \config_n$. By the induction hypothesis, either $\config_i$ is poised or it takes a step (where $i \in \set{1, \ldots, n}$). If any $\config_i$ takes a step then $\config$ takes a step. Otherwise, all $\config_i$ are poised, so $\config$ is poised.

  \item For the single channel case, we know $\config = \config', \proc c P$. By inversion, $\typeRecD {\ctx_0} \emptyset P c A$ and $\providesCtx {\config'} {\ctx_0}$. By the induction hypothesis, either $\config'$ takes a step, in which case $\config$ takes a step and we are done, or $\config'$ is poised. Assume $\config'$ is poised.

  Define $\pred{\typeRecD{\ctx'}{\emptyset}{P}{c}{A'}} :=$ if $\providesCtx {\config'} {\ctx'}$ then either $\proc c P$ is poised or $\config$ can take a step. We proceed by induction on the typing derivation.
    \begin{description}
      \item[Case $\id:$] $P$ has the form $\tfwd c d$ and $\config$ steps by $\irb{id}$.
      \item[Case $\cut:$] $P$ has the form $\tspawn d {Q_d} {P'_d}$ and $\config$ steps by $\irb{cut}$.

      \item[Case $\terminate\Right, \tensor\Right, \internal\Right, \lolli\Right, \external\Right :$] $\proc c P$ is poised.

      \item[Case $\terminate\Left, \tensor\Left, \internal\Left, \lolli\Left, \external\Left :$] The proofs for all these cases follow the same structure, so we will only present $\terminate\Left$. We know $P = \twait d P'$ for some $P'$ and $d : \terminate \in \ctx'$. By inversion on $\providesCtx {\config'} {\ctx'}$, we get $\typeRecD {\ctx''} \emptyset Q d \terminate$ and $\proc d Q \in \config'$ for some $\ctx''$ and $Q$. $\proc d Q$ is poised since $\config'$ is poised by the outer induction hypothesis, so \cref{inversion-poised-eq} gives $Q = \tclose d$. Thus, $\config$ steps by $\irb{one}$.

      \item[Case $\rec :$] $P$ has the form $\tapp {\parens{\trec t {\tvector y} P'}} {\tvector z}$ and $\config$ steps by $\irb{rec}$.

      \item[Case $\typeeq\Right :$] Follows immediately by the induction hypothesis.
      \item[Case $\typeeq\Left :$] $\ctx' = \ctx'', d : D$ and $\typeRecD{\ctx'', d : D'}{\emptyset}{P}{c}{A'}$ for some $D'$ where $D \typeeq D'$. We then know $\ctx' \typeeq \ctx'', d : D'$, thus, the result follows from \cref{typeeq-configuration} and the induction hypothesis.
    \end{description}

    Finally, we get either $\poised{\proc c P}$ or $\config$ steps from $\pred{\typeRecD {\ctx_0} \emptyset P c A}$ and $\providesCtx{\config'}{\ctx_0}$. In the former case, $\config$ is poised since $\config'$ is poised from before, and in the latter case we are immediately done.
\end{itemize}
\end{proof}

\subsection{Type Preservation}

Preservation is a bit more tedious to prove.

\begin{theorem}[Preservation]
\label{base:preservation}
If $\providesCtx \config \ctx$ and $\steps{\config}{\config'}$ then $\providesCtx {\config'} \ctx$.
\end{theorem}

\begin{proof}
  By \cref{framing}, it suffices to consider the subtree which types the root of reduction. So, assume $\steps{\config_1}{\config_2}$ and $\provides {\config_1} c A$ where $c$ is the root of $\steps{\config_1}{\config_2}$. We need to show $\provides {\config_2} c A$.

  The proof is by simultaneous case analysis on $\steps{\config_1}{\config_2}$ and induction on the typing derivation of the root process, followed by induction on the typing derivation of the provider in cases where there is communication. We need induction rather than simple inversion due to $\typeeq\Right$ and $\typeeq\Left$, which change types on the right and the left (respectively) without exposing the structure of the process.

  By inversion, $\config_1 = \parens{\config_1^c, \proc c P}$, $\typeRecD \ctx \emptyset P c A$, and $\providesCtx {\config_1^c} \ctx$. Define $\pred{\typeRecD {\ctx_c} \emptyset P c {A'}} :=$ if $\providesCtx {\config_1^c} {\ctx_c}$ then $\provides {\config_2} c {A'}$. We proceed by induction.
  %
  \begin{itemize}
    %% Base cases
    \item $\predC{\infer[\id]{\typeRecD {d : A'} \emptyset {\tfwd c d} c {A'}}%
            {\phantom{id}}} :$
      \par Then, $\config_2 = \subst c d {\config_1^c}$. Note that $c \not\in \dom{\parens{\config_1^c}}$ (since well-formed contexts do not have duplicate channel labels), so $\provides {\config_1^c} d {A'}$ implies $\provides {\config_2} c {A'}$ by substitution.

    \item $\predC{\infer[\cut]{\typeRecD {\ctx_c, \ctx_c'} \emptyset {\tspawn d {P_d'} { Q_d }} c {A'}}%
            { \DD : \typeRecD {\ctx_c} \emptyset {P_d'} d D
            & \EE : \typeRecD {\ctx_c', d : D} \emptyset c {Q_d} {A'}
            } } :$
     \par Then, $\config_2 = \config_1^c, \proc a {P_a'}, \proc c {Q_a}$ where $a$ is fresh. By inversion on $\providesCtx {\config_1^c} {\parens{\ctx_c, \ctx_c'}}$, there are $\config_1^1, \config_1^2$ such that $\config_1^c = \parens{\config_1^1, \config_1^2}$, $\providesCtx {\config_1^1} {\ctx_c}$ and $\providesCtx {\config_1^2} {\ctx_c'}$. Applying $\confOne$ on $\DD$ and $\providesCtx {\config_1^1} {\ctx_c}$ gives $\provides {\config_1^1, \proc a {P_a'}} a D$ (through suitable substitution). Then,  $\providesCtx {\config_1^2, \config_1^1, \proc a {P_a'}} {\parens{\ctx_c', a : D}}$ by $\confN$. Finally, $\EE$ and $\confOne$ implies $\provides {\config_2} c {A'}$.

      \item $\terminate\Right, \tensor\Right, \internal\Right, \lolli\Right, \external\Right :$ Impossible since $\proc c P$ is the client.

      %% Nested Induction
      \item $\terminate\Left, \tensor\Left, \internal\Left, \lolli\Left, \external\Left :$ In all cases, $\ctx_c = \ctx_c', d : D$ for some $D$. Inversion on $\providesCtx {\config_1^c} {\ctx_c}$ gives $\config_1^c = \config_1^{c'}, \config_1^d, \proc d Q$ such that $\typeRecD {\ctx'} \emptyset Q d D$, $\providesCtx {\config_1^d} {\ctx'}$, and $\providesCtx {\config_1^{c'}} {\ctx_c'}$.

        Define $\predQ{\typeRecD {\ctx_d} \emptyset Q d {D'}} :=$ if $\providesCtx {\config_1^d} {\ctx_d}$ and $D' \typeeq D$ then $\provides {\config_2} c {A'}$. Note that $\predQ{\typeRecD {\ctx'} \emptyset Q d D}$ will give the final result once we prove $\mathcal{Q}$, which we do by induction.
        \begin{itemize}
          %%% Base cases
          \item $\predQC{\infer[\terminate\Right]{\typeRecD {\emptyCtx} \emptyset {\tclose d} d \terminate}%
            {} }$,
            $\predC{\infer[\terminate\Left]{\typeRecD {\ctx_c', d : \terminate} \emptyset {\twait d {P'}} c {A'}}%
              { \DD : \typeRecD {\ctx_c'} \emptyset {P'} c {A'}
              } } :$

            \par Then, $\config_2 = \config_1^{c'}, \config_1^d, \proc c {P'} = \config_1^{c'}, \proc c {P'}$ (last equality is by inversion on $\providesCtx {\config_1^d} {\emptyCtx}$). $\confOne$ on $\providesCtx {\config_1^{c'}} {\ctx_c'}$ and $\DD$ gives the desired result.

          \item $\predQC{\infer[\tensor\Right]{\typeRecD {\ctx_d, \ctx_d'} \emptyset {\tsend d e {R_e} {Q'}} d {D_1' \tensor D_2'}}%
            { \DD : \typeRecD {\ctx_d} \emptyset {R_e} e {D_1'}
            & \EE : \typeRecD {\ctx_d'} \emptyset {Q'} d {D_2'}
            } }$,
            \par $\predC{\infer[\tensor\Left]{\typeRecD {\ctx_c', d : D_1 \tensor D_2} \emptyset {\trecv x d {P_x'}} c {A'}}%
              { \FF : \typeRecD {\ctx_c', x : D_1, d : D_2} \emptyset {P_x'} c {A'}
              } } :$
            \par Then $\config_2 = \config_1^{c'}, \config_1^d, \proc a {R_a}, \proc d {Q'}, \proc c {P_a'}$ where $a$ is fresh. Inversion on $D_1' \tensor D_2' \typeeq D_1 \tensor D_2$ gives $D_1' \typeeq D_1$ and $D_2' \typeeq D_2$. There are $\config_1^1, \config_1^2$ such that $\providesCtx {\config_1^1} {\ctx_d}$ and $\providesCtx {\config_1^2} {\ctx_d'}$ by inversion on $\providesCtx {\config_1^d} {\parens{\ctx_d, \ctx_d'}}$.

            $\confOne$ on $\providesCtx {\config_1^1} {\ctx_d}$ and $\DD$ with $\typeeq\Right$ gives $\provides {\config_1^1, \proc a {R_a}} a {D_1}$. Similarly, $\confOne$ on $\providesCtx {\config_1^2} {\ctx_d'}$ and $\EE$ with $\typeeq\Right$ gives $\provides {\config_1^2, \proc d {Q'}} d {D_2}$. Finally, $\confOne$ using the previous two derivations, $\providesCtx {\config_1^{c'}} {\ctx_c'}$, and $\FF$ with gives the desired result.

          \item $\internal\Right, \lolli\Right, \external\Right :$ Similar to above.

          %%% Inductive Cases
          \item $\predQC{\infer[\typeeq\Right]{\typeRecD {\ctx_d} \emptyset Q d {D'}}%
            { \DD : \typeRecD {\ctx_d} \emptyset Q d {D''}
            & \EE : D'' \typeeq D'
            } } :$
            \par $D'' \typeeq D$ by transitivity of $\typeeq$ (\cref{eq-is-equivalence}), so we can immediately apply the induction hypothesis on $\DD$.

          \item $\predQC{\infer[\typeeq\Right]{\typeRecD {\ctx_d, e : E} \emptyset Q d {D'}}%
            { \DD : \typeRecD {\ctx_d, e : E'} \emptyset Q d {D'}
            & \EE : E \typeeq E'
            } } :$
            \par $\ctx_d, e : E \typeeq \ctx_d, e : E'$ using $\EE$, so $\providesCtx {\config_1^d} {\parens{\ctx_d, e : E'}}$ by \cref{typeeq-configuration}. Thus, we can apply the induction hypothesis on $\DD$, which gives the desired result.

          %%% Impossible cases
          \item $\id, \cut, \terminate\Left, \tensor\Left, \internal\Left, \lolli\Left, \external\Right, \rec :$ Not applicable since we know the form of $Q$ by the outer induction and inversion over $\steps{\config_1}{\config_2}$.
        \end{itemize}

    \item $\predC{\infer[\rec]{\typeRecD {\ctx_c} \emptyset P c {A'}}%
            { \DD : \typeRecD {\ctx_c} \emptyset P c {A''}
            } } :$
      \par Follows from a suitable substitution lemma for process variables. We omit the details since it is quite standard.

    %% Induction Step
    \item $\predC{\infer[\typeeq\Right]{\typeRecD {\ctx_c} \emptyset P c {A'}}%
            { \DD : \typeRecD {\ctx_c} \emptyset P c {A''}
            & \EE : A'' \typeeq A'
            } } :$
      \par Induction hypothesis on $\DD$ gives $\provides {\config_2} c {A''}$. The result follows from \cref{typeeq-configuration} using $\EE$.

    \item $\predC{\infer[\typeeq\Left]{\typeRecD {\ctx_c, d : D} \emptyset P c {A'}}%
            { \DD : \typeRecD {\ctx_c, d : D'} \emptyset P c {A'}
            & \EE : D \typeeq D'
            } } :$
      \par $D \typeeq D'$ implies $\ctx, d : D \typeeq \ctx, d : D'$ by definition. \Cref{typeeq-configuration} gives $\providesCtx {\config_1^c} {\parens{\ctx, d : D'}}$, which means we can apply the induction hypothesis on $\DD$ to get the desired result.
  \end{itemize}

  Left out cases are impossible since $\steps{\config_1}{\config_2}$ and the structure of the root process must match up. Finally, $\pred{\typeRecD \ctx \emptyset P c A}$ gives the desired result.
\end{proof}
