
\section{Process Expressions}
\label{base:process-expressions}

Within this framework, proof terms correspond to processes. For example, cut, written $\tspawn{c}{P_c}{Q_c}$, denotes a form of process composition where the client spawns off a helper process ($P_c$) with which it can communicate from then on. The intuition is formalized in the typing rule:
$$ \infer[\cut]{ \typeD {\ctx, \ctx'} {\tspawn{c}{P_c}{Q_c}} {d} {D} }
    { \typeD {\ctx} {P_c} {c} {A}
    & \typeD {\ctx', c : A} {Q_c} {d} {D}
    }
$$

The rest of the process expressions are summarized below, with the sending construct
followed by the matching receiving construct. Discussion of other typing rules is given in \cref{base:type-assignment}.

\begin{center}
\begin{tabular}{l c l l}
  $P, Q, R$ & ::= & $\tspawn{x}{P_x}{Q_x}$     & cut (spawn) \\
            & $|$ & $\tfwd c d$                & id (forward) \\
            & $|$ & $\tclose c \mid \twait c P$  & $\terminate$ \\
            & $|$ & $\tsend{c}{y}{P_y}{Q} \mid \trecv{x}{c}{R_x}$ & $A \tensor B,$ $A \lolli B$ \\
            & $|$ & $\tselect{c}{}{P} \mid \tcase c {\tbranches Q i}$  & $\externals A I,$ $\internals A I$ \\
            & $|$ & $\tsendVal{c}{M}{Q} \mid \trecvVal{n}{c}{R_n}$ & $A \sendVal B,$ $A \recvVal B$
\end{tabular}
\end{center}

Note that cut, $\mathtt{send}$, and $\mathtt{recv}$ bind the spawned, sent, and received channel names, which means these are identified up to alpha conversion. We denote the usual capture avoiding simultaneous substitution of channels $\tvector c$ for $\tvector x$ in $P$ by $\subst {\tvector c} {\tvector x} P$ where $\tvector c$ and $\tvector x$ are ordered sequences of equal length.


\subsection{An Example Process: Process Level Naturals}

Let us consider an example program to get more intuition about the system. We will use process level natural numbers, $\nat$, as a running example. Examples are given using concrete syntax, which closely follows the abstract syntax presented above with some syntactic sugar. Briefly, channel names start with a \texttt{`}. Type declarations have the form \texttt{type <type name> = <definition>} and process declarations have the form \texttt{<offered channel> <- <process name> = <definition>}. Process declarations can refer to the offered channel, and both kinds of declarations can use the declared name recursively.%
\footnote{Recursive types and processes are introduced formally in \cref{base:recursion}.}
All declarations are considered mutually recursive. We define $\tsendChannel c d P = \tsend c x {\tfwd x d} P$ and $\tfwd c P = \tspawn x P {\parens{\tfwd c x}}$ as syntactic sugar. If a process declaration has a type $A_1 \lolli A_2 \lolli \ldots \lolli A_n \lolli B$, then it can be applied to $n$ channels of the corresponding types, which is translated into a spawn followed by $n$ sends. The full grammar is given in \cref{concrete-syntax}.

With that out of the way, we can give our first example. First, we define the interface:

\begin{lstlisting}[language=krill, style=custom]
  type Nat = +{zero : 1, succ : Nat}
\end{lstlisting}

The interface states that a process level natural number is an internal choice of either zero or a successor of another natural. Next, we define two simple processes that implement the interface:

\begin{minipage}{.48\textwidth}
\begin{lstlisting}[language=krill, style=custom]
  z : Nat
  `c <- z =
    `c.zero;
    close `c
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}{.48\textwidth}
\begin{lstlisting}[language=krill, style=custom]
  s : Nat -o Nat
  `c <- s `d =
    `c.succ;
    `c <- `d
\end{lstlisting}
\end{minipage}

\texttt{z} simply sends the label \texttt{zero} along the channel \texttt{`c} (which it provides) and terminates, whereas \texttt{s} send the label \texttt{succ} and delegates to \texttt{`d}. Here is a slightly more complicated example that uses recursion:

\begin{lstlisting}[language=krill, style=custom]
  double : Nat -o Nat
  `c <- double `d =
    case `d of
      zero -> `c.zero; wait `d; close `c
      succ -> `c.succ; `c.succ; `c <- double `d
\end{lstlisting}


\subsection{Type Assignment for Processes}
\label{base:type-assignment}

The typing rules for other constructs are derived from linear logic by decorating derivations with proof terms. The rules are given in \cref{session-assignment}. Note that we allow unused branches case expressions for $\internal\Left$ and $\external\Right$, which makes width subtyping easier (discussed in \cref{base:subtyping}).

\begin{rules}[session-assignment]{Type assignment to processes}
  % id and cut
  \infer[\id]{ \typeD {c : A} {\tfwd{d}{c}} {d} {A} }
    {}
  \and \infer[\cut]{ \typeD {\ctx, \ctx'} {\tspawn{c}{P_c}{Q_c}} {d} {D} }
    { \typeD {\ctx} {P_c} {c} {A}
    & \typeD {\ctx', c : A} {Q_c} {d} {D}
    }
  % Terminate
  \and \infer[\terminate\Right]{\typeD{\emptyCtx}{\tclose c}{c}{\terminate}}
    {}
  \and \infer[\terminate\Left]{\typeD{\ctx, c : \terminate}{\twait c P}{d}{A}}
    {\typeDJ{P}{d}{A}}
  % Tensor
  \and \infer[\tensor\Right]{\typeD{\ctx, \ctx'}{\tsend{c}{d}{P_d}{Q}}{c}{A \tensor B}}
    { \typeD{\ctx}{P}{d}{A}
    & \typeD{\ctx'}{Q}{c}{B}
    }
  \and \infer[\tensor\Left]{ \typeD{\ctx, c : A \tensor B}{\trecv{d}{c}{P_d}}{e}{E} }
    { \typeD{\ctx, d : A, c : B}{P_d}{e}{E} }
  % Internal choice
  \and \infer[\internal\Right]{\typeDJ { \tselect{c}{i}{P} } {c} {\internals{A}{I}} }
    { i \in I
    & \typeDJ{P}{c}{A_i}
    }
  \and \infer[\internal\Left]{ \typeD { \ctx, c : \internals{A}{I} } { \tcase{c}{\tbranches{P}{J}} } {d} {D} }
   { I \subseteq J
   & \typeD{\ctx, c : A_k}{P_k}{d}{D}~\text{for}~k\in I
   }
  % Implication
  \and \infer[\lolli\Right]{ \typeD{\ctx}{\trecv{d}{c}{P_d}}{c}{A \lolli B} }
    { \typeD{\ctx, d : A}{P_d}{c}{B} }
  \and \infer[\lolli\Left]{\typeD{\ctx, \ctx', c : A \lolli B}{ \tsend{c}{d}{P_d}{Q} } {e}{E}}
    { \typeD{\ctx}{P_d}{d}{A}
    & \typeD{\ctx', c : B}{Q}{e}{E}
    }
  % External choice
  \and \infer[\external\Right]{ \typeDJ { \tcase{c}{\tbranches{P}{I}} } {c} {\externals{A}{J}} }
   { J \subseteq I
   & \typeDJ{P_k}{c}{A_k}~\text{for}~k\in J
   }
  \and \infer[\external\Left]{\typeD{\ctx, c : \externals{A}{I}} { \tselect{c}{i}{P} } {d} {D}}
    { i \in I
    & \typeD{\ctx, c : A_i}{P}{d}{D}
    }
\end{rules}

As usual, we identify bound channels up to alpha conversion. Free channels are subject to consistent renaming across a sequent by substitution.
%We can change free channel names while preserving typing as formalized below:
% \begin{lemma}[Channel Substitution]
%   \label{channel-substitution-pre}
%   If $\typeDJ P c A$ and $\tvector z \cap \free {\parens P} = \emptyset$ then $\typeD {\subst {\tvector z} {\tvector y} \ctx} {\subst {\tvector z} {\tvector y} P} {\subst {\tvector z} {\tvector y} c} A$.
% \end{lemma}

% We delay the proof until after we introduce recursive types since the typing-derivation will be changed slightly.


